Supporting information for ``How environmental drivers of synchrony interact''

### 2022-01-13 Paper/Paper.Rnw
continued working on Methods

@

@@

One of the classically theorized \citep{Moran1953} and now widely empirically validated mechanisms of 
synchrony, called the Moran effect, operates when spatially synchronized environmental signals, such
as temperature or precipitation, help drive population dynamics and thereby induce synchrony in
populations. Two recent papers have documented that Moran effects of distinct 
environmental drivers can, in fact, interact, either syergistically or antagonistically,
so that total population synchrony can be either greater or less than if the distinct Moran 
drivers operated independently \citep{Sheppard2019,CastoraniSheppardKelp}. Effects can be strong.
\cite{Sheppard2019} showed that long-timescale ($>4$y period) synchrony in a chlorophyll density index 
in the seas around the United Kingdom is substantially augmented by interactions between environmental 
drivers of synchrony and drivers linked with consumption by a copepod predator.
\cite{CastoraniSheppardKelp} showed that nutrient dynamics and wave disturbance, two Moran drivers
of synchrony in giant kelp populations on the California coast, interact either synergistically 
or destructively, depending on which portion of the California coast is examined and on 
timescale of analysis. 
Both of these studies argued that interactions between Moran effects may be a general feature of many 
systems, because of the large number of interrelated factors driving population fluctuations in most 
spatially extended systems. The main purpose of this study is to establish a general theory of the mechanism
of interacting Moran effects and to use examples and applications of the theory to provide 
ecological intuition for how the mechanism of interacting Moran effects works. Our goals are
distinct from those of \cite{Kendall2000}, who examined interactions between Moran and dispersal
causes of synchrony. We relate our work to that study in the Discussion.

@@

Thus our specific goals for thus study are the following. First, (G1) we will elaborate a general mathematical 
theory of interacting Moran effects which allows a detailed understanding of how the mechanism works. Second, 
(G2) we will consider theoretical case studies, which emerge as important special cases of our general theory, and 
which illuminate the intuition behind how Moran effects may interact in real systems. Third, (G3) we will 
apply our theory to populations of giant kelp off the California coast. Whereas \cite{CastoraniSheppardKelp}
already carried out a detailed analysis of interacting Moran effects in California kelp populations and their importance 
for kelp ecology, we instead use a simplified subset of the available kelp data, and our analysis is instead intended 
to illuminate the inner workings of our theory and of the general mechanisms we describe. In the Discussion, we 
extend the arguments of \cite{Sheppard2019} and \cite{CastoraniSheppardKelp} that interactions between Moran 
effects are likely to be common, and hence that our general theory seems likely to pertain to numerous 
ecosystems. Overall, this study serves to introduce a general theory of Moran interactions and to conceptually
illuminate a newly observed but potentially quite common mechanism of spatiotemporal population dynamics.

@@

\noindent We now describe how we pursue goal G2 of the Introduction, to develop 
theoretical case studies that illuminate the intuition of interacting Moran effects. 
We develop case studies A-C. For all cases, the model time step was assumed to be one quarter, i.e., 
four model time steps per year. This makes no real mathematical difference, but was done to 
facilitate later comparisons with results based on kelp data, which were available with a
quarterly sampling frequency. 
For case study A,  the environmental variable 
$\epsilon^{(1)}$ is assumed to have a simple positive effect on populations, but 
lagged by one model time step, i.e.,
$p_0^{(1)}=0$, $p_1^{(1)}>0$, and $p_i^{(1)}=0$ for $i>1$. 
For case study B, which generalizes A, $\epsilon^{(1)}$ is again assumed to have a simple 
positive effect on populations,
but now lagged by $n$ time steps, so that $p_n^{(1)}>0$ while $p_i^{(1)}=0$ for $i \neq n$.
For both A and B, the effects of $\epsilon^{(2)}$
are assumed to be un-lagged, i.e., $p_0^{(2)} \neq 0$ and $p_i^{(2)}=0$ for $i>0$.
For case study C, the effects of $\epsilon^{(1)}$ are lagged by $n_1$ and the 
effects of $\epsilon^{(2)}$ are lagged by $n_2$ time steps, i.e.,
$p_{n_1}^{(1)} \neq 0$ while $p_i^{(1)}=0$ for $i \neq n_1$, and 
$p_{n_2}^{(2)} \neq 0$ while $p_i^{(2)}=0$ for $i \neq n_2$.
The noise process
$(\epsilon^{(1)},\epsilon^{(2)})$ is assumed to be a Gaussian white-noise process 
for both A and B. So 
the random variables $(\epsilon^{(1)}(t),\epsilon^{(2)}(t))$ are independent and 
identically distributed (iid) for distinct times, $t$. Noise may be correlated across space,
and the components of $\epsilon^{(1)}(t)$ may be correlated with those of $\epsilon^{(2)}(t)$.
For C, the noise processes $\epsilon^{(1)}$ and $\epsilon^{(2)}$ are
each assumed to exhibit spatially synchronous periodic oscillations of period one year, i.e.,
four model time steps, but with different phases. Peaks in $\epsilon^{(1)}$ are assumed to 
either lead or lag (we consider two model scenarios) peaks in $\epsilon^{(2)}$ by one quarter.
Such a scenario could be realized by annually 
periodic environmental fluctuations sampled quarterly,
e.g., kelp wave action in central California peaks annually in the winter, whereas surface-water
nitrate concentrations also fluctuate with period one year, but peak in the spring, a delay of one quarter
compared to the wave peak. 
Details of the noise processes are specified in SI section \ref{SM-sect:examples_cross_sync}.
For all case studies,
the autoregressive order of population dynamics is assumed, for simplicity, to be $1$, i.e, $b_1 \neq 0$ 
but $b_i=0$ for $i>1$.

%Tentative broad outline
%0) above stuff about the case studies 
%1) Para from Tom on how remote sensing data are gathered and coverted into what he handed to me. Brief, mostly 
%citing other works where it is laid out in much more detail. 
%2) Para from Tom on how he got to the wave and NO3 data
%3) Para summarizing the data cleaning I did
%Comment) Actually, might want to send the abive 3 paras to sup mat, and just have one para describing the final format of the
%data, especially if you make this paper also be about plankton. But do that later if you do it.
%5) Para about the linear modelling - this is how certain parts of the main theoretical equation were estimated
%6) Para about spectral estimates. Brief, cites appropriate parts of Brillinger, uses sup mat if needed. This is how certain
%other parts of the main theoretical equation were estimated. 
%Comment) Somehow need to get across what was done - different parts of main equation were examined to explore the different
%aspects of how the mechanism works for the focal taxa. Elaborate as appropriate.

%Para 1 from Tom
A process was previously developed (REFS) using remote sensing to provide quarterly kelp biomass and kelp canopy area estimates,
from quarter 1 of 1987 to quarter 4 of 2019, at $437$ locations along the California coast, from slightly north of Monterey Bay
to San Diego. We very briefly describe the process, refering to our earlier published works for details.
%Need to mention: The spacing and extent of each site; where the data can be downloaded (other than in our git repo, though I'll
%a mention of that, too); lots of references as necessary; probably need slightly more formal statements of what "kelp biomass" 
%"kelp canopy area" mean in these contextx; anything else important. Need bibliographic data for cited papers.

%Para 2 from Tom
Another process was previously developed (REFS) to provide mean nitrate concentration and maximum wave height data
for the same quarters and locations. We again briefly describe the process, citing earlier works for details.

%I asked Tom for these paragraphs by email on 2021 08 05

%Need:
%1) Some kind of map of the locations and display of some time series for the sup mat, to give a sense of the data. I think Max will
%have made one of these for his paper with Lawrence, so maybe I could get him to make another one, same or appropriately modified?

@@

%Note that we get differences in intereaction effects between central and southern CA because we 
%have lagged effects of nitrate on observed kelp density in central CA, but unlagged effects in southern
%CA (to see this, look at the regression coefficients computed to get the models used for the
%spectral analysis, which were saved). Careful inspection of my ms figures show this is the reason
%for the positive versus slightly negative interaction effects seen in central v. souther CA. 
%But I spke with Tom and Max and Kyle about this by email around 1/3/2022. They agreed that 
%probably it has something to do with the fact that kelp is ripped out by waves in central
%CA, so it takes a quarter to recover enough to be detectable from space (we can only see the 
%canopy from space, on the surface of the water). In southern CA, waves rip out kelp like that
%much less. So if kelp is always around and usually reaches the surface in southern CA, nitrate
%effects can be observed more quickly. There ought to be a Discussion paragraph on this, basiaclly
%saying things may be more complex that our analyses reveals with regard to some of the differences 
%we observed between central and southern CA. We didn't explore this because our goal was to illustrate 
%our theory, and our work does demonstrate the importance of interaction effects for the synchrony of kelp
%as observed from space. But future work could look at this more, certainly, perhaps using state-space models
%which can separate biological from observation processes.

@@

>>>>

resloc<-"../Results/Kelp_DataAfterAllCleaning/"
quarters<-readRDS(paste0(resloc,"Quarters_CleanedFinal.Rds"))

resloc<-"../Results/Kelp_SpectralDecompResults/"
SBlocstouse<-readRDS(paste0(resloc,"SBlocstouse.Rds"))
CC1locstouse<-readRDS(paste0(resloc,"CC1locstouse.Rds"))
CC2locstouse<-readRDS(paste0(resloc,"CC2locstouse.Rds"))
@

@@

One of the classically theorized \citep{Moran1953} and now widely empirically validated mechanisms of 
synchrony, called the Moran effect, operates when spatially synchronized environmental signals, such
as temperature or precipitation, help drive population dynamics and thereby induce synchrony in
populations. Two recent papers have documented that Moran effects of distinct 
environmental drivers can, in fact, interact, either syergistically or antagonistically,
so that total population synchrony can be either greater or less than if the distinct Moran 
drivers operated independently \citep{Sheppard2019,CastoraniSheppardKelp}. Effects can be strong.
\cite{Sheppard2019} showed that long-timescale ($>4$y period) synchrony in a chlorophyll density index 
in the seas around the United Kingdom is substantially augmented by interactions between environmental 
drivers of synchrony and drivers linked with consumption by a copepod predator.
\cite{CastoraniSheppardKelp} showed that nutrient dynamics and wave disturbance, two Moran drivers
of synchrony in giant kelp populations on the California (CA) coast, interact either synergistically 
or destructively, depending on which portion of the CA coast is examined and on 
timescale of analysis. 
Both of these studies argued that interactions between Moran effects may be a general feature of many 
systems, because of the large number of interrelated factors driving population fluctuations in most 
spatially extended systems. The main purpose of this study is to establish a general theory of the mechanism
of interacting Moran effects and to use examples and applications of the theory to provide 
ecological intuition for how the mechanism of interacting Moran effects works. Our goals are
distinct from those of \cite{Kendall2000}, who examined interactions between Moran and dispersal
causes of synchrony. We relate our work to that study in the Discussion.

@@

Thus our specific goals for thus study are the following. First, (G1) we will elaborate a general mathematical 
theory of interacting Moran effects which allows a detailed understanding of how the mechanism works. Second, 
(G2) we will consider theoretical case studies, which emerge as important special cases of our general theory, and 
which illuminate the intuition behind how Moran effects may interact in real systems. Third, (G3) we will 
apply our theory to populations of giant kelp off the CA coast. Whereas \cite{CastoraniSheppardKelp}
already carried out a detailed analysis of interacting Moran effects in CA kelp populations and their importance 
for kelp ecology, we instead use a simplified subset of the available kelp data, and our analysis is instead intended 
to illuminate the inner workings of our theory and of the general mechanisms we describe. In the Discussion, we 
extend the arguments of \cite{Sheppard2019} and \cite{CastoraniSheppardKelp} that interactions between Moran 
effects are likely to be common, and hence that our general theory seems likely to pertain to numerous 
ecosystems. Overall, this study introduces a general theory of Moran interactions and uses it to conceptually
illuminate this newly observed but potentially quite common mechanism of spatiotemporal population dynamics.

@@

%Theoretical case studies
\noindent We now describe how we pursue goal G2 of the Introduction, to develop 
theoretical case studies that illuminate the intuition of interacting Moran effects. 
We develop case studies A-C. For all cases, the model time step was assumed to be one quarter, i.e., 
four model time steps per year. This makes no real mathematical difference, but was done to 
facilitate later comparisons with results based on kelp data, which were available with a
quarterly sampling frequency. 
For case study A,  the environmental variable 
$\epsilon^{(1)}$ is assumed to have a simple positive effect on populations, but 
lagged by one model time step, i.e.,
$p_0^{(1)}=0$, $p_1^{(1)}>0$, and $p_i^{(1)}=0$ for $i>1$. 
For case study B, which generalizes A, $\epsilon^{(1)}$ is again assumed to have a simple 
positive effect on populations,
but now lagged by $n$ time steps, so that $p_n^{(1)}>0$ while $p_i^{(1)}=0$ for $i \neq n$.
For both A and B, the effects of $\epsilon^{(2)}$
are assumed to be un-lagged, i.e., $p_0^{(2)} \neq 0$ and $p_i^{(2)}=0$ for $i>0$.
For case study C, the effects of $\epsilon^{(1)}$ are lagged by $n_1$ and the 
effects of $\epsilon^{(2)}$ are lagged by $n_2$ time steps, i.e.,
$p_{n_1}^{(1)} \neq 0$ while $p_i^{(1)}=0$ for $i \neq n_1$, and 
$p_{n_2}^{(2)} \neq 0$ while $p_i^{(2)}=0$ for $i \neq n_2$.
The noise process
$(\epsilon^{(1)},\epsilon^{(2)})$ is assumed to be a Gaussian white-noise process 
for both A and B. So 
the random variables $(\epsilon^{(1)}(t),\epsilon^{(2)}(t))$ are independent and 
identically distributed (iid) for distinct times, $t$. Noise may be correlated across space,
and the components of $\epsilon^{(1)}(t)$ may be correlated with those of $\epsilon^{(2)}(t)$.
For C, the noise processes $\epsilon^{(1)}$ and $\epsilon^{(2)}$ are
each assumed to exhibit spatially synchronous periodic oscillations of period one year, i.e.,
four model time steps, but with different phases. Peaks in $\epsilon^{(1)}$ are assumed to 
either lead or lag (we consider two model scenarios) peaks in $\epsilon^{(2)}$ by one quarter.
Such a scenario could be realized by annually 
periodic environmental fluctuations sampled quarterly,
e.g., kelp wave action in central CA peaks annually in the winter, whereas surface-water
nitrate concentrations also fluctuate with period one year, but peak in the spring, a delay of one quarter
compared to the wave peak. 
Details of the noise processes are specified in SI section \ref{SM-sect:examples_cross_sync}.
For all case studies,
the autoregressive order of population dynamics is assumed, for simplicity, to be $1$, i.e, $b_1 \neq 0$ 
but $b_i=0$ for $i>1$.

%Description of data after cleaning
The core of this paper is the new theory, but to help illustrate that theory, we also apply
it to an exceptional dataset on giant kelp (\emph{Macrocyctis pyrifera}) dynamics off the CA coast, and 
associated environmental measurements. 
The data are based on a subset of those used by \cite{CastoraniSheppardKelp} and
\cite{Walter2021}, and data details are given in those papers and in SI section \ref{SM-sect:data}. We here
summarize the format of the data after a preparation and cleaning process was implemented. 
\cite{CastoraniSheppardKelp} and \cite{Walter2021} also provide an introduction to 
kelp ecology, and information on why giant kelp is an excellent species 
for studies of synchrony. After preparation, kelp data consisted of
$\Sexpr{length(SBlocstouse)+length(CC1locstouse)+length(CC2locstouse)}$ quarterly kelp abundance time series
from locations along the CA coast, each time series spanning from quarter $\Sexpr{quarters[1,3]}$ 
of $\Sexpr{quarters[1,2]}$ to quarter $\Sexpr{quarters[dim(quarters)[1],3]}$ of 
$\Sexpr{quarters[dim(quarters)[1],2]}$.   
Time series were grouped into three regions which were analyzed separately: a more northerly
central CA group of $\Sexpr{length(CC1locstouse)}$ locations (called CCA1); a more southerly central CA group of 
$\Sexpr{length(CC2locstouse)}$ locations (CCA2); and a group of $\Sexpr{length(SBlocstouse)}$ locations 
from southern CA, close to Santa Barbara
(SB; see Fig. \ref{fig:map}). Each kelp measurement is an estimate of mean quarterly kelp biomass per unit 
useable habitat in a 500m stretch of coastline. We used coastline segments where kelp was persistent
through essentially all of the $\Sexpr{quarters[1,2]}$-$\Sexpr{quarters[dim(quarters)[1],2]}$ 
period (SI section \ref{SM-sect:data}). We also had estimates
of maximum wave height and mean nitrate concentration for each quarter and location; both waves
and nutrients are known to influence kelp population dynamics and synchrony 
\citep{Cavanaugh2013,CastoraniSheppardKelp}.

%Tentative broad outline
%2) Para about the linear modelling - this is how certain parts of the main theoretical equation were estimated,
%about the lags we used, and justification for them etc. Might also need to cite a sup mat section. Maybe 
%we need one single sup mat section called "supplementary methods" or "details of methods" or something. Actually
%let's have two of these, one for data, one for stats
%3) Para about spectral estimates. Brief, cites appropriate parts of Brillinger, uses sup mat if needed. This is how certain
%other parts of the main theoretical equation were estimated. Basically it's just about how to do spectral
%estimates of spectra and cross spectra, and then rest follows from the main theoretical equation.
%4) Somehow need to get across what was done - different parts of main equation were examined to explore the different
%aspects of how the mechanism works for the focal taxon. Elaborate as appropriate.

Data for the project are archived at X.
%***DAN: Need to get the data location from Jon, I think he posted it, right? Maybe he did not put it all i
%one place. Check with him. At the end of the Max/Lawrence paper, several data locations are listed.
%I have an email out to Jon and Max.
All computations were done in R version 3.4.4 on an ordinary laptop which was running Ubuntu Linux
16.04. Complete codes and data are at \url{https://github.com/reumandc/InteractingMoranEffects.git}.

@@

%Note that we get differences in intereaction effects between central and southern CA because we 
%have lagged effects of nitrate on observed kelp density in central CA, but unlagged effects in southern
%CA (to see this, look at the regression coefficients computed to get the models used for the
%spectral analysis, which were saved). Careful inspection of my ms figures show this is the reason
%for the positive versus slightly negative interaction effects seen in central v. souther CA. 
%But I spke with Tom and Max and Kyle about this by email around 1/3/2022. They agreed that 
%probably it has something to do with the fact that kelp is ripped out by waves in central
%CA, so it takes a quarter to recover enough to be detectable from space (we can only see the 
%canopy from space, on the surface of the water). In southern CA, waves rip out kelp like that
%much less. So if kelp is always around and usually reaches the surface in southern CA, nitrate
%effects can be observed more quickly. There ought to be a Discussion paragraph on this, basiaclly
%saying things may be more complex that our analyses reveals with regard to some of the differences 
%we observed between central and southern CA. We didn't explore this because our goal was to illustrate 
%our theory, and our work does demonstrate the importance of interaction effects for the synchrony of kelp
%as observed from space. But future work could look at this more, certainly, perhaps using state-space models
%which can separate biological from observation processes.   
%   Tom also told me in a meeting zoom call that in central CA growth of new fronds will be slower because 
%there are not other fronds (left) on the same plant to provide a sugar subsidy to new fronds. In central CA,
%basically all fronds from the same plant are ripped out.

%Need a paragraph (already mentioned above) on how a frequency/timescale-specific approach
%is necessary for all this. Need to think more about it first, i.e., how would you attempt to 
%do this without such an approach, and how far would you get? Or maybe that is not quite the 
%right at it (though you should try thinking that through and see what you get), maybe another
%possibly better way would be to see that everything we list here can depend on timescale.
%You can get a non-timescale-specific approach just by integrating our main equation across
%frequencies. However, that will obscure the fact that you can get strong positive interaction 
%effects on some timescales and strong negative ones on others. What would you get in our 
%theoretical case studies if you integrated and would it reveal what's really going on? What 
%are the implications of timescale structure for practical situations where populations and
%environmental variables are measured over a certain typical duration (i.e, short time series)?

@@

\begin{figure}
\caption{To fill.}\label{fig:map}
\end{figure}

### 2022-01-13 Paper/SupMat.Rnw
continued working on Methods

@

@@

\section{Details of data and data cleaning}

%***DAN: This should include specification of the regions CC1, CC2, SB.
%You have to get some info from Tom about the nature of the spatial sampling, and the process leading up to what he gave me.
%You have to then describe the stages of cleaning I did.

>>>>

resloc<-"../Results/Kelp_DataAfterSuperBasicCleaning/"
kelp_SB<-readRDS(paste0(resloc,"Kelp_Quarterly_CleanedSuperBasic.Rds"))
resloc<-"../Results/Kelp_DataAfterMoreInvolvedCleaning/"
kelp_MI<-readRDS(paste0(resloc,"Kelp_Quarterly_CleanedMoreInvolved.Rds"))
@

@@

\section{Details of data and data cleaning}\label{sect:data}

Floating surface canopy of kelp was measured by remote sensing, with details and
validation reported by \cite{Cavanaugh2011} and \cite{Bell2020, Bell2021}. Multispectral Landsat
imagery of $30$m resolution was used. Quarterly mean biomass values and spatial extent of canopy  
within $500$m coastline segments (henceforth called ``loactions'')
were computed by averaging. Quarters were: Q1, January-March; Q2, April-June; 
Q3, July-September; Q4, October-December. 
Data were thusly available from $\Sexpr{dim(kelp_SB)[1]}$ locations spanning 
approximately from Monterey Bay to San Diego. 
We then chose locations for analysis for which kelp was ``persistent'', in this 
context meaning absent or missing data for no more than three years; this was to focus on 
population dynamics of kelp rather than extinction-recolonization dynamics, which occur in some 
locations and which can relate to sediment dynamics on the ocean floor and other factors
not currently of primary interest. Of the original $\Sexpr{dim(kelp_SB)[1]}$ locations,
$\Sexpr{dim(kelp_MI)[1]}$ ($\Sexpr{100*round(dim(kelp_MI)[1]/dim(kelp_SB)[1],2)}$\%) 
were deemed persistent according to this criterion. Missing values in the remaining 
time series were then filled with seasonal medians. The biomass and spatial extent time series
for each location were then used together to produce one time series estimating biomass
density per unit useable habitat at the site, as follows: every value of 
the biomass time series was divided by the $90^{\text{th}}$ quantile of the distribution
of the spatial extent time series at the site. The $90^{\text{th}}$ quantile was used instead
of the maximum value of the spatial extent time series because maximum statistics have high variance. 
This procedure served to normalize total biomass by an estimate
of total useable habitat at a site. Kelp establish on rocky substrate, and areas with sandy or
sedimented substrate are not useable kelp habitat. All time series were then linearly detrended,
and only those time series were utilized which were located within the three regions pictured
in Fig. \ref{MT-fig:map} in the main text. Regions were selected to represent distinct dynamical
patterns known to occur in central and southern CA, and also to have another the same number
of time series per selected region.

Wave nutrient data were also available for each quarter at each location. Wave data were derived 
by applying a swell propagation model with bouy-based wave measurements and a validation 
techniques \citep{Wingeart2001,Hansen2009,Bell2015,OReilly2016}. Data on nitrates 
were produced \citep{Snyder2020} by exploiting an established empirical 
relationship between surface nitrate concentration
and sea surface temperature (SST) in coastal CA \citep{Zimmerman1984,Palacios2013,Snyder2020}
and SST data derived from Advanced Very High Resolution Radiometer satellite imagery. 
Wave and nutrient time series in each location were again linearly detrended before use.

\section{Details of statistical methods}
